Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-15 manuscript/draft.Rmd
Anna comments

---
title: "Best practices for forecasting changes in biodiversity: an example using breeding birds"
author:
- David J. Harris
- Shawn D. Taylor
- Ethan P. White
bibliography: refs.bib
csl: csl.csl
geometry: left=3.5cm, right=3.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
fontsize: 12pt
header-includes:
    - \usepackage{times}
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
output:
  pdf_document: default
  html_document: default
---

@@

Biodiversity forecasts are important for conservation, management, and
evaluating how well current models characterize natural systems. While
biodiversity forecasts are common in the literature, there is little information
available on how well these forecasts work. Most biodiversity forecasts are not
evaluated in terms of their ability to predict future diversity, fail to account
for uncertainty, and do not use time-series data that captures the actual
dynamics being studied. We compiled a list of ten best practices for forecasting
biodiversity, and used hindcasting to evaluate six different modeling approaches
for predicting breeding bird species richness. Each method was evaluated at `r
numbers$N_predict_sites` sites distributed throughout the continental United
States. Forecasts were evaluated for up to a decade after the final training
observations. While each model could explain most of the variance in richness,
none of them consistently outperformed a baseline model that simply predicted
constant richness at each site. In particular, we found no evidence that current
methods (such as species distribution models) can turn spatial data into useful
temporal predictions about biodiversity. This result may call into question the
validity of many high-profile forecasts from the last 15 years. Most of the best
practices implemented in this study (e.g., evaluating confidence intervals as
well as point estimates) directly influence the outcome of the forecasts; some
even led to large changes in the relative performance of different modeling
approaches. To facilitate the rapid improvement of biodiversity forecasts, we
emphasize the value of specific best practices in making and evaluating
different forecasting methods.

@@

Forecasting the future state of ecological systems is increasingly important for
planning, managing, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001; @houlahan2017;
@dietze2017]. Forecasting is biodiversity is especially important, due to its
central role in conservation planning and its sensitivity to anthropogenic
effects [@cardinale2012; @diaz2015; @tilman2017]. Anticipating potential changes
in biodiversity is crucial for large scale management and conservation, and for
addressing debates regarding whether local scale diversity is declining and may
continue to decline in response to anthropogenic influences [@diaz2015;
@dornelas2014].

@@

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity, little is known about how effectively we can anticipate their
dynamics. This is due in part to the long time scales over which many ecological
forecasts are applied [and the resulting difficulty in assessing whether the
predicted changes occurred @dietze2016]. What we do know comes from a small
number of hindcasting studies, where models are built using data on species
occurrence and richness from the past and evaluated on their ability to predict
contemporary patterns [e.g., @algar2009; @distler2015]. These studies are a
valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation [@clark2001,
@dietze2017]; and 3) conducting meaningful evaluations of the forecasts by
hindcasting, archiving short-term forecasts, and comparing forecasts to
baselines to determine whether the forecasts are more accurate than simply
assuming the system is basically static [@perretti2013].

@@

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 30 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. This 30 year
time-scale has only recently become possible for large-scale time-series based
forecasting and allows us to model and assess changes a decade or more into the
future in the presence of shifts in environmental conditions on par with
predicted climate change. We compare traditional distribution modeling based
approaches to spatial models of species richness, time-series methods, and
simple baselines (Table 1). All forecasts model uncertainty and observation
error, are evaluated across different time lags using hindcasting, and are
publicly archived to allow future assessment. We discuss the implications of
these practices for our understanding of, and confidence in, the resulting
forecasts, and how we can continue to build on these approaches improve
ecological forecasting in the future.

@@

@@

Bird species richness was obtained from the North American Breeding Bird Survey
(BBS) [@pardieck2017] using the Data Retriever software [@morris2013] and
rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
all nocturnal, cepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. All data from routes that did not meet BBS acceptable criteria were
also excluded.

Environmental data included a combination of elevation, bioclimatic variables
and a remotely sensed vegetation index (the normalized difference vegetation
index; NDVI), all of which are known to influence richness and distribution in
the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions [@harris2015]
for each year in the dataset: mean diurnal range, isothermality, max temperature
of the warmest month, mean temperature of the wettest quarter, mean temperature
of the driest quarter, precipitation seasonality, precipitation of the wettest
quarter, precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], with an 8 km resolution was
obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
period 1981-2013. Average summer (May, June, July) and winter (December,
January, Feburary) NDVI values were used as predictors. Elevation was from the
SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. All environmental variables were aggregated to the average
values within a 40 km radius (the length of a BBS route) of the starting point
of each BBS route.

We made long term forecasts from 2014-2050 using CMIP5 multi-model ensemble
dataset as the source for climatic variables. Precipitation and temperature from
37 downscaled model runs [@brekke2013, see Table S1] using the RCP6.0 scenario
were averaged together to create a single ensemble used to calculate the
bioclimatic variables for North America. For NDVI we used the per-site average
values from 2000-2013 as a simple forecast. For observer effects (see below) each
site was set to have zero observer bias.

We used observed richness values from 1982 (the first complete year of NDVI
data) to 2003 for training the models, and data from 2004 to 2013 was used for
testing their performance. We only used BBS routes from the continental United
States (i.e. routes where PRISM data was available), and restricted the analysis
to routes that were sampled during 70% of the years in the training period
(i.e., routes with at least 16 annual observations). The resulting dataset
included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique sites,
and included `r numbers$N_species` species. Site-level richness varied from `r
numbers$richness_summary$min` to `r numbers$richness_summary$max` with an
average richness of `r numbers$richness_summary$mean` species.

@@

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan]. Because observer and site are strongly related
(observers tend to repeatedly sample the same site), site was also included as a
random effect to ensure that inferred deviations were actually observer-related
(as opposed to being related to the sites that a given observer happened to
see). The resulting model partitions the variance in observed richness values
into site-level variance, observer-level variance, and residual variance (e.g.
variation within a site from year to year; Figure 3). The site-level estimates
can also be used directly as the "average" baseline model (see below). The
estimated observer effects can be subtracted from the richness values for a
particular observer to provide an estimate of how many species would have been
found by a "typical" observer. To incorporate uncertainty in these "corrected"
richness values into the forecasting models we collected 500 Monte Carlo samples
from the model's posterior distribution, and fit each of the downstream models
with each of the Monte Carlo samples.

@@

### Site-level models

We fit three sets of models that were fit to each site independently, with no
environmental information. These models were fit to each BBS route both on the
raw richness values and on the residuals from our observer model. When
correcting for observer effects, we averaged across 500 models that were fit
separately to the 500 Monte Carlo estimates of the observer effects, to account
for our uncertainty in the true values of those effects. All of these models use
a Gaussian error distribution (rather than a count distribution) for the reasons
discussed below .

**Baseline models.** We used two simple models as baselines for comparison to
the fits of more complex models. These baselines treated site-level richness
observations either as uncorrelated noise around a site-level constant (the
"average" model) or as an autoregressive model with a single year of history
(the "naive" model) [@hyndman2014]. Predictions from the "average" model are
centered on the average richness observed during training, and the confidence
intervals are narrow and constant-width. The "naive" model, in contrast, 
predicts that future observations will be similar to the final observed value
(e.g. the value observed in `r timeframe$last_train_year`), and the confidence
intervals expand rapidly as the predictions extend farther into the future.

@@

### Continental-scale models

>>>>

---
title: "Best practices for forecasting changes in biodiversity: an example using breeding birds"
author:
- David J. Harris
- Shawn D. Taylor
- Ethan P. White
bibliography: refs.bib
csl: csl.csl
geometry: left=3.5cm, right=3.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
fontsize: 12pt
header-includes:
    - \usepackage{times}
    - \usepackage{setspace}
    - \usepackage{booktabs}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
output:
  pdf_document: default
  html_document: default
---

@@

Biodiversity forecasts are important for conservation, management, and
evaluating how well current models characterize natural systems. While
biodiversity forecasts are common, there is little information available on how
well these forecasts work. Most biodiversity forecasts are not evaluated in
terms of their ability to predict future diversity, fail to account for
uncertainty, and do not use time-series data that captures the actual dynamics
being studied. We compiled a list of ten best practices for forecasting
biodiversity, and used hindcasting to evaluate six different modeling approaches
for predicting breeding bird species richness. Each method was evaluated at `r
numbers$N_predict_sites` sites distributed throughout the continental United
States. Forecasts were evaluated for up to a decade after the final training
observations. While each model could explain most of the variance in richness,
none of them consistently outperformed a baseline model that simply predicted
constant richness at each site. In particular, we found no evidence that current
methods (such as species distribution models) can successfully turn spatial data
into useful temporal predictions about biodiversity. This result may call into
question the validity of many high-profile forecasts from the last 15 years.
Most of the best practices implemented in this study (e.g., evaluating
confidence intervals as well as point estimates) directly influence the outcome
of the forecasts; some even led to large changes in the relative performance of
different modeling approaches. To facilitate the rapid improvement of
biodiversity forecasts, we emphasize the value of specific best practices in
making and evaluating different forecasting methods.

@@

Forecasting the future state of ecological systems is increasingly important for
planning and management, and also for quantitatively evaluating how well our
models capture the key ecological processes governing these systems [@clark2001;
@houlahan2017; @dietze2017]. Forecasts regarding biodiversity are especially
important, due to biodiversity's central role in conservation planning and its
sensitivity to anthropogenic effects [@cardinale2012; @diaz2015; @tilman2017].
High-profile studies forecasting large biodiversity declines over the coming
decades have played a large role in shaping ecologists' priorities [as well as
those of policymakers; @ipcc2014], but it is inherently difficult to evaluate
such long-term predictions before the projected biodiversity declines have
occurred.

@@

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity [[refs]], little is known about how effectively we can
anticipate these dynamics. This is due in part to the long time scales over
which many ecological forecasts are applied [and the resulting difficulty in
assessing whether the predicted changes occurred; @dietze2016]. What we do know
comes from a small number of hindcasting studies, where models are built using
data on species occurrence and richness from the past and evaluated on their
ability to predict contemporary patterns [e.g., @algar2009; @distler2015]. These
studies are a valuable first step, but lack several components that are
important for developing forecasting models with high predictive accuracy, and
for understanding how well different methods can predict the future. The
components an effective forecasting and evaluation strategy (Box 1) broadly
involve: 1) expanding the use of data to include biological and environmental
time-series [@tredennick2016]; 2) accounting for uncertainty in observations,
processes, model choice, and forecast evaluation [@clark2001; @dietze2017]; and
3) conducting meaningful evaluations of the forecasts by hindcasting, archiving
short-term forecasts, and comparing forecasts to baselines to determine whether
the forecasts are more accurate than simply assuming the system is basically
static [@perretti2013].

@@

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of sites located throughout North America, while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 30 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. Datasets that
span a time scale of 30 years or more have only recently become available for
large-scale time-series based forecasting. A dataset of this size allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and two simple baselines that predict
constant richness for each site, on average (Table 1). All of our forecasting
models account for uncertainty and observation error, are evaluated across
different time lags using hindcasting, and are publicly archived to allow future
assessment. We discuss the implications of these practices for our understanding
of, and confidence in, the resulting forecasts, and how we can continue to build
on these approaches to improve ecological forecasting in the future.

@@

We evaluated 6 types of forecasting models (Table 1) at two time-scales. In the
first one, we divided our [30] years of data into [22] years of training data
and retained [9] years of data for evaluation via hindcasting. Second, we used
the full data set for training and made predictions through the year 2050. For
both time scales, we evaluated versions of each model with and without
correcting for observer effects, as described below.

@@

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever software
[@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data was
filtered to exclude all nocturnal, cepuscular, and aquatic species [since these
species are not well sampled by BBS methods; @hurlbert2005], as well as
unidentified species, and hybrids. All data from routes that did not meet BBS
acceptable criteria were also excluded.

**Past environmental data.** Environmental data included a combination of
elevation, bioclimatic variables and a remotely sensed vegetation index (the
normalized difference vegetation index; NDVI), all of which are known to
influence richness and distribution in the BBS data [[ref]]. For each year in
the dataset, we used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions
[@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
month, mean temperature of the wettest quarter, mean temperature of the driest
quarter, precipitation seasonality, precipitation of the wettest quarter, and
precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
dataset with an 8 km resolution [@pinzon2014] and was available from the time
period 1981-2013. Average summer (May, June, July) and winter (December,
January, Feburary) NDVI values were used as predictors. Elevation was from the
SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. Because BBS routes are 40-km transects rather than point counts,
we used the average value of each environmental variable within a 40 km radius
of each BBS route's starting point.

**Future environmental projections.** We made long term forecasts from 2014-2050
using the CMIP5 multi-model ensemble dataset as the source for climatic
variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
runs [@brekke2013, see Table S1] using the RCP6.0 scenario were averaged
together to create a single ensemble used to calculate the bioclimatic variables
for North America. For NDVI we used the per-site average values from 2000-2013
as a simple forecast. For observer effects (see below) each site was set to have
zero observer bias.

**[[xxx]].** We used observed richness values from 1982 (the first complete year
of NDVI data) to 2003 for training the models, and data from 2004 to 2013 was
used for testing their performance. We only used BBS routes from the continental
United States (i.e. routes where PRISM data was available), and restricted the
analysis to routes that were sampled during 70% of the years in the training
period (i.e., routes with at least 16 annual observations). The resulting
dataset included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique
sites, and included `r numbers$N_species` species. Site-level richness varied
from `r numbers$richness_summary$min` to `r numbers$richness_summary$max` with
an average richness of `r numbers$richness_summary$mean` species.

@@

For each forecasting approach, we trained two versions of the corresponding
model: one with corrections for differences among observers, and one without.
Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan]. Because observer and site are strongly related
(observers tend to repeatedly sample the same site), site was also included as a
random effect to ensure that inferred deviations were actually observer-related
(as opposed to being related to the sites that a given observer happened to
see). The resulting model partitions the variance in observed richness values
into site-level variance, observer-level variance, and residual variance (e.g.
variation within a site from year to year; Figure 3). The site-level estimates
can also be used directly as the "average" baseline model (see below). The
estimated observer effects can be subtracted from the richness values for a
particular observer to provide an estimate of how many species would have been
found by a "typical" observer. To incorporate uncertainty in these "corrected"
richness values into the forecasting models we collected 500 Monte Carlo samples
from the model's posterior distribution, and fit each of the downstream models
with each of the Monte Carlo samples. Each Monte Carlo sample represented a
different possible set of observer-level and site-level random effect values
across the full 30-year dataset.

@@

### Models: site-level models

We fit three sets of models that were fit to each site separately, with no
environmental information (Table 1). These models were fit to each BBS route
twice: once using the residuals from the observer model, and once using the raw
richness values. When correcting for observer effects, we averaged across 500
models that were fit separately to the 500 Monte Carlo estimates of the observer
effects, to account for our uncertainty in the true values of those effects. All
of these models use a Gaussian error distribution (rather than a count
distribution) for the reasons discussed below (see "Model evaluation").

**Baseline models.** We used two simple baseline models as a basis for
comparison with the more complex models (Figure 2A). These baselines treated
site-level richness observations either as uncorrelated noise around a
site-level constant (the "average" model) or as an autoregressive model with a
single year of history (the "naive" model) [@hyndman2014]. Predictions from the
"average" model are centered on the average richness observed during training,
and the confidence intervals are narrow and constant-width. The "naive" model,
in contrast, predicts that future observations will be similar to the final
observed value (e.g. the value observed in `r timeframe$last_train_year`), and
the confidence intervals expand rapidly as the predictions extend farther into
the future. Both models' richness predictions are centered on a constant value,
so neither model can anticipate any trends in richness or any responses to
future environmental changes.

@@

### Models: continental-scale models
